Sampling-based Motion Planning for Tracking Evolution of Dynamic Tunnels in Molecular Dynamics Simulations

### 2017-10-31 paper.tex
from mrs gitlab

Therefore, researchers have started to focus on molecular dynamics (MD) simulations to study the behavior of individual tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole,jurcik2016accelerated}.
%tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
Molecular dynamics is represented by a sequence of time frames (molecule snapshots), and the tunnels need to be detected through these frames.
%With the increasing size of simulations the correctness of predicting the most biochemically relevant tunnel increases as well.
The existing solutions are based exclusively on Voronoi diagrams, suffering from several problems.

First, to search for the correspondence between tunnels in individual snapshots, these solutions are using different clustering methods, which are generally very time and memory consuming.
This also limits the maximum number of frames which can be analyzed.
In such cases, the biochemists have to select a subset of the whole simulation and perform the analysis only for this selection.
Such analysis however gives only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted.

Second, the basic parameter determining if a tunnel will be detected is the bottleneck radius, i.e., the size of the narrowest part of the tunnel. 
This parameter is set to a specific value and even a slight decrease of the bottleneck radius below this threshold causes that the tunnel is not detected.
In consequence, the tunnels detected within an MD simulation are not continuous -- there are snapshots where the given tunnel is completely missing and appears again after some time.
In fact this does not correspond to the real situation. 
The tunnel void space does not disappear in these snapshots, it only does not connect the active site with the protein surface by a sufficiently wide path.
To be able to trace the void space also in these snapshots, we introduce so called {\it dynamic tunnel}.
By this term we denote a void space between the protein active site and its surface, whose position and length is traced over time.
The dynamic tunnel does not need to be connected with the active site or the surface all the time which gives the users the information about the void space also in snapshots without traditional tunnels.
In consequence, the user gets much better idea about the behavior of the void space which can be used by a ligand to reach or leave the active site.

In our proposed solution to these problems, we utilize the algorithms from sampling-based motion planning.
We have recently proposed a novel method for tunnel detection using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT) method~\cite{vonasek2017tunnel}.
RRT builds a tree of collision-free configurations of a spherical probe moving inside the protein. 
To consider the molecular dynamics, the tree is continuously pruned in each frame.
In comparison with the existing approaches~\cite{Petrek20071357,petrek2006caver}, this RRT-based approach~\cite{vonasek2016application} 
handles the dynamics without any need to cluster and match Voronoi diagrams between consecutive frames, which leads to results much faster and with less memory consumption.

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{0.1}
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{cc}
\includegraphics[width=0.17\textwidth]{fig/motiv1} &
\includegraphics[width=0.19\textwidth]{fig/motiv2lab} \\
Protein 1CQW & Active site \\
\multicolumn{2}{c}{\includegraphics[width=0.19\textwidth]{fig/motiv3}} \\
\multicolumn{2}{c}{Static tunnels }
\end{tabular}
\begin{tabular}{cc}
\includegraphics[width=0.26\textwidth]{fig/dt-f5} &
\includegraphics[width=0.26\textwidth]{fig/dt-f6} \\
Frame $n$ & Frame $n+1$ \\
\includegraphics[width=0.26\textwidth]{fig/dt-f7} &
\includegraphics[width=0.26\textwidth]{fig/dt-f8} \\
Frame $n+2$ & Frame $n+3$ \\
\end{tabular}
}
\caption{\label{fig::motiv}
Example of static tunnel detection in DhaA haloalkane dehalogenase and the example of a dynamic tunnel that evolves through four consecutive frames.
In the frame $n$, the active site (red sphere) is not fully connected with a protein surface, but only to an internal cavity.
The cavity disconnects from the active site in the frame $n+1$ and remains disconnected also in the next frame $n+2$.
In the fourth frame $n+3$, the cavity connects with the protein surface.
}
\end{figure}

@@

Alternatively, the tunnel detection in dynamic proteins can be formulated as path planning in 4D (i.e., 3D $\times$ time) configuration space assuming a spherical probe.
This space can be searched using sampling-based motion planning methods like Probabilistic Roadmaps (PRM)~\cite{kavrakiPRM} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} and their modifications.
In sampling-based motion planning, the configuration space is randomly sampled and the samples are classified as free or non-free using collision detection.
The collision detection is performed between the geometry of the robot (i.e., the probe) and the environment (the atoms of the protein).
The collision-free samples are stored into a roadmap which is searched using standard graph-search methods for a path.
The path in the roadmap then relates to the motion in the workspace.

@@

In~\cite{guieysse2008structure,cortes2010simulating}, RRT is used to compute exit pathways for flexible ligand leading from an active site to the surface of the protein.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.
In~\cite{cortesMD}, the pathway is found for the opposite direction, from the active site towards the protein surface.
The methods~\cite{guieysse2008structure,cortesMD} do not compute all tunnels explicitly, rather a single pathway.

@@

A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples
in the difficult areas.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling e.g.
around medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}, or even by shifting
uniformly generated samples towards the medial axis~\cite{amatoOBPRM} or its estimation~\cite{hollemanMAPRM}.
Another approach is to estimate difficult regions e.g. based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
Generating random samples in difficult regions however does not ensure the construction of a better tree in the case of RRT-based methods.
As RRT builds the tree incrementally, the tree can connect to a region with increased probability of sampling only if the growth
of the tree is not blocked by obstacles.
Instead of increasing probability of sampling in all regions, e.g. along the whole medial axis, the tree has to guided through the space.
For example, the random samples are generated from multiple regions, that are switched according to the progress of the tree.
In~\cite{kardossRRTKK}, the samples are generated along several waypoints that are located close to narrow passages.
A more general approach is to define a guiding path in the workspace and sequentially increase the probability of sampling along
the path~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}.
%The guiding path can be computed e.g. using medial axis of the worskpace~\cite{denny2014marrt,denny2016dynamic}.

@@

Molecular dynamics simulation is represented by a sequence of consecutive snapshots (frames), where each snapshot describes the position of all protein atoms in a specific time.
Proteins are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
The geometry of the protein in the $i-$th frame is represented by the union of all spheres $\SS^i \subset \R^3$.
The void space is detected using the spherical probe of radius $\probe$. 
All possible configurations $q=(x,y,z)$ (positions) of the probe in the $i$-th frame form the configuration space $\C^i$.
The probe is represented by a sphere $\Sprobe(q) \subset \R^3$ at a configuration $q$.
The void space (the collision-free region) $\CF^i \subseteq \C^i$ is formed by configurations, where the $\Sprobe$ can be placed without any collision with atoms, i.e.,  $\CF^i = \{q \in \C^i | \Sprobe(q) \cap \SS^i = \emptyset\}$.
For each configuration $q\in\C^i$ we assume that its largest collision-free radius $r(q) \in \mathbf{R}, r(q)\ge 0$ can be computed, e.g., 
using collision detection.
The outer space  $\CG^i \in \C^i$ of the protein is defined by configurations, where a large sphere can be placed without any collision,
i.e., $\CG^i=\{q\in \C^i| \Sgprobe(q) \cap \SS^i = \emptyset \}$, where $\Sgprobe$ is the sphere of radius $\gprobe > \probe$ 
placed at the configuration $q$.
To determine whether two configurations $q_1, q_2 \in \CF^i$ can be connected, a line segment from $q_1$ to $q_2$ is constructed
and discretized to several points, using $\varepsilon$ discretization step and the collision detection is performed at all the points.
The two configurations can be connected if all intermediate points on the line are collision-free.

The main purpose of the tunnel detection is the identification of tunnels, either static or dynamic, that connect the active site with the protein surface.
We assume that the position (configuration) of the active site $\qinit \in \CF^i$ in each snapshot is known in advance or it is known that the probe 
cannot be placed at the active site.

The approximated WVD diagram of the atoms is computed by representing each atom using 12 balls of equal radii and computing the 
ordinary Voronoi diagram.
Computing WVD this way is numerically more stable and it is used also in other related tools~\cite{caver3,yaffe2008}.
The vertices of this WVD are filtered out if their distance to an atom is less than the probe radius $\probe$ or if their
distance to the nearest atom is larger than $\gprobe$.
These filtered Voronoi vertices are stored in the set $\VV^i$.

The atoms located on the surface of the molecule can be identified using \mbox{$\alpha$-shape}. 
Let $\SSA^i \subset \SS^i$ denote a set of spheres representing these atoms in snapshot $i$.
%The distance between a configuration $q \in \C^i$ and its nearest surface atom is referred to as $\dists(q)$.
Since all configurations describe the position of the probe in 3D space, {\textcolor{red} {all distance evaluations between two configurations, between a configuration and a sphere}} are measured using the 3D Euclidean metric, which is denoted as $\dist(\cdot,\cdot)$ in the rest of the paper.

@@

To detect the void space, the collision-free region $\CF^i$ of the $i$-th frame is searched and represented as a tree structure of the collision-free configurations.
The detection starts from a given active site $\qinit \in \CF^i$ or from a set of already detected collision-free positions of the probe.
The second case is a result of the void space detection in the previous frame of the molecular dynamics.
To maintain the connectivity between the void space from the previous frame and the newly discovered one, the void space
has to be explored incrementally without adding non-reachable regions, such as cavities.
Due to this reason, the void space cannot be automatically initialized using all Voronoi vertices $\VV^i$, though they are collision-free
at the radius $\probe$, because the Voronoi vertices can be also placed in the cavities.
The required incremental detection of the void space can be achieved by the exploration principle of the RRT planner, as the RRT incrementally extends a tree of collision-free configurations.
The main loop of the algorithm for the void space detection is listed in Alg.~\ref{alg::main}.

@@

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
{\small
\caption{\label{alg::main}voidSpaceDetection}
\KwIn{Initial configuration tree $\T$,
    initial set of blocking spheres $\SB'$,
    position of active site $\qinit\in\CF^i$,
}
\KwData{goal-probe radius $\gprobe$,
    probe radius $\probe$, 
    number of iterations~$\Imax$, 
    phase--switching threshold~$\alpha$,
    prob. $\gb$~of sampling randomly from $\C^i$,
}
\KwOut{Configuration tree $\T$,
    list of tunnel exit-points $P$
}
\hrule
$\SB = \SB'$; //blocking atoms added during phase~1 \\
$P = \emptyset$; // list of tunnel exit-points \\
$phase=1$\;
$\VV^i = \VV^i \cup \{ \qinit \}$; // add active site position in order to reach it by the tree\\
updateCollisionDetection($\SB$)\;       
\For{$iteration = 1:\Imax$}{
    \If(// Voronoi-based sample --- Section~\ref{sec::vbg}){$phase=1$}{
        $\VVA^i$ = updateActiveVoronoiVertices($\T, \VV^i$)\;
        $\VVA^i$ = sort vertices according to the distance to the tree\; \nllabel{alg::main::d}
        \ForEach{$v \in \VVA^i$}{
            $\qnear$ = nearestNode($\T, v$)\;
            $\T,P', \SB'$ = expandTree($\T, \qnear, v$)\;
            $P = P \cup P'$\; \nllabel{alg::main::g1}
            $\SB = \SB \cup \SB'$\;
            updateCollisionDetection($\SB$)\; \nllabel{alg::main::a} \nllabel{alg::main::e}
        }
    }
    \If{$phase=1$}{ \nllabel{alg::main::f}
        $v =$ random point from $\VVA^i$\;
        $\qrand =$ random point $\sim N(v,r_v)$\; \nllabel{alg::main::gauss}
    }
    \If{$phase=2$ {\bf or} rand(0,1) $< p_c$}{
        $\qrand$ = random sample from $\C^i$\; \nllabel{alg::main::g}
    }

@@

Besides the exploration of the void space, it is necessary to detect the exit points as well.
The exit points are configurations from $\CG^i$, i.e., they are collision-free at the radius $\gprobe$ and they define the end of the tunnel.
The detection of the exit points requires that the tree exits the protein.
A premature exit from the protein may lead to quick filling of the outer part of the protein and eventually even to entering the protein from outside. 
Such an exploration would lead to the detection of a smaller number of tunnels.
An example as depicted in Fig.~\ref{fig::blocking1}).

To ensure that the tree explores primarily the internal void space while allowing its growth outside the protein for the purpose of exit point detection, the algorithm is divided into two phases.
In the first phase, the growth of the tree outside the protein is blocked using a set of {\sl blocking spheres} $\SB$.
The blocking spheres are placed at configurations that are located close to the surface atoms $\SSA^i$.
Before a configuration $q$ is added to the tree, the distance between $q$ and the surface atoms is computed.
If the distance is smaller than a predefined threshold $\dts$, $q$ can be assumed to be close to the surface and a new blocking sphere of radius $\gprobe$ is added to $\SB$.

In the second phase, the blocking spheres are not used and they are removed from the collision detection (Alg.~\ref{alg::main}, line~\ref{alg::main::b}).
This allows the tree to explore also the void space outside the protein, which is necessary to find configurations representing the tunnel exit points.
The outer space is more likely reached from the nodes that could not be expanded in the first phase due to the blocking spheres.
As these nodes are located close to the surface atoms, this situation does not require many iterations to expand these nodes.
The second phase of the algorithm should therefore be shorter than the first one.
In Alg.~\ref{alg::main}, the second phase starts after $\alpha$ fraction of all iterations.

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{1pt}
\begin{tabular}{cccc}
\includegraphics[width=0.24\textwidth]{fig/blockingb1} &
\includegraphics[width=0.24\textwidth]{fig/blockingb2} & 
\includegraphics[width=0.24\textwidth]{fig/blockingb3} & 
\includegraphics[width=0.24\textwidth]{fig/blockingb4} \\
a & b & c & d \\
\end{tabular}  
}
\caption{\label{fig::blocking1}
    Example of the tree growth without the blocking spheres. 
    The initial configuration $\qinit$ (active site) is represented by the black circle.
    The tree starts to grow inside the protein (a) but it can grow outside as well (b).
    In region denoted by A, the tree grows back to the protein (c).
    Finally, two tunnels are detected (with the exit points {\textcolor{red} {$P_1$ and $P_2$)}}, but any tunnel is found in the region $A$ (d).
}
\end{figure}

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{1pt}
\begin{tabular}{cccc}
\includegraphics[width=0.24\textwidth]{fig/blockingb5} &
\includegraphics[width=0.24\textwidth]{fig/blockingb6} &
\includegraphics[width=0.24\textwidth]{fig/blockingb7} & 
\includegraphics[width=0.24\textwidth]{fig/blockingb9} \\
a & b & c & d \\
\includegraphics[width=0.24\textwidth]{fig/blockingb10a} &
\includegraphics[width=0.24\textwidth]{fig/blockingb10b} & 
\includegraphics[width=0.24\textwidth]{fig/blockingb10c} & 
\includegraphics[width=0.24\textwidth]{fig/blockingb10d} \\
e & f & g & h \\
\end{tabular}
}
\caption{\label{fig::blocking2}
The effect of the blocking spheres. 
The surface atoms are depicted in brown.
After the tree reaches a surface atom, a new blocking sphere (red large circle) is added to the collision detection (a).
This forces the tree to explore only the internal void space (b,c).
In the second phase of the algorithm, the blocking spheres are removed and the tree can finally grow outside the protein in order
to detect exit points (d).
Figures e--h depict four detected tunnels.
}
\end{figure}

@@

The generation of the random samples directly influences the performance of the RRT-based planners.
In classic RRT, the random samples are drawn uniformly in the whole configuration space, which leads to the well known narrow passage problem.
The void space can be considered as a narrow passage, as the movement of the probe is constrained by many atoms.
The limitation is even higher for larger probes.
To boost the growth of the tree inside the narrow void space, the random samples can be generated around places that are believed to be a part
of the collision-free region $\CF^i$.
This region is approximated by the Voronoi vertices $\VV^i$ and the samples can be generated around them.

However, the probability of sampling cannot be increased around all Voronoi vertices at once, because the growth of the tree towards the Voronoi vertices may be blocked by the obstacles (atoms)~\cite{vonasekphd}.
It is necessary to guide the growth of the tree by increasing the probability of sampling only around the Voronoi vertices that can be reached by the tree with a higher probability.
Such Voronoi vertices have to be located close to the tree.
Therefore, a subset $\VVA^i \subseteq \VV^i$ of the Voronoi vertices that are close to the tree is maintained (Alg.~\ref{alg::update}).
A Voronoi vertex $v \in \VV^i$ is considered as active if the distance to its nearest node in the tree is less than $\da$, but larger than
$\db$.
The upper value $\da$ ensures that the active Voronoi vertices are close to the boundary of the configuration tree, whilst the lower value $\db$ prevents to activate the Voronoi vertices that have already been approached by the tree.
After the active Voronoi vertices are updated, {\textcolor{red} {they are sorted according to their distance to the tree attempts to extend towards them}} (Alg.~\ref{alg::main}, line~\ref{alg::main::d}--\ref{alg::main::e}).

>>>>

Therefore, researchers have started to focus on molecular dynamics (MD) simulations to study the behavior of individual tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole,jurcik2016accelerated}.
%tunnels over time~\cite{yaffe2008,caver3,sehnal2013mole}.
Molecular dynamics is represented by a sequence of time frames (molecule snapshots), and the tunnels need to be detected through these frames.
%With the increasing size of simulations the correctness of predicting the most biochemically relevant tunnel increases as well.
The existing solutions are based exclusively on Voronoi diagrams and they suffer from several problems.

First, to determine for the correspondence between tunnels in individual snapshots, these solutions employ various clustering methods, which are generally very time and memory consuming.
This also limits the maximum number of frames which can be analyzed.
In such cases, the biochemists have to select a subset of the whole simulation and perform the analysis only for this selection.
Such analysis however gives only a rough idea of the tunnel behavior, as important parts of the simulation can be easily omitted.

Second, the basic parameter determining if a tunnel will be detected is the bottleneck radius, i.e., the size of the narrowest part of the tunnel. 
This parameter is set to a specific value and even a slight decrease of the bottleneck radius below this threshold causes that the tunnel is not detected.
In consequence, the tunnels detected within an MD simulation are not continuous -- there are snapshots where the given tunnel is completely missing and appears again after some time.
In fact this does not correspond to the real situation. 
The tunnel void space does not disappear in these snapshots, it only does not connect the active site with the protein surface by a sufficiently wide path.
To be able to trace the void space also in these snapshots, we introduce so called {\it dynamic tunnel}.
By this term we denote a void space between the protein active site and its surface, whose position and length is traced over time.
The dynamic tunnel does not need to be connected with the active site or the surface all the time.
By detecting these tunnel, users gain more information about the void space also in snapshots without traditional tunnels.

In our proposed solution to these problems, we utilize the algorithms from sampling-based motion planning.
We have recently proposed a novel method for tunnel detection using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT) algorithm~\cite{vonasek2017tunnel}.
RRT builds a tree of collision-free configurations of a spherical probe moving inside the protein. 
To consider the molecular dynamics, the tree is continuously pruned in each frame.
In comparison with the existing approaches~\cite{Petrek20071357,petrek2006caver}, this RRT-based approach~\cite{vonasek2016application} 
handles the dynamics without any need to cluster and match Voronoi diagrams between consecutive frames, which 
leads to faster and memory less demanding computation.

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{0.1}
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{cc}
\includegraphics[width=0.17\textwidth]{fig/motiv1} &
\includegraphics[width=0.19\textwidth]{fig/motiv2lab} \\
Protein 1CQW & Active site \\
\multicolumn{2}{c}{\includegraphics[width=0.19\textwidth]{fig/motiv3}} \\
\multicolumn{2}{c}{Static tunnels }
\end{tabular}
\begin{tabular}{cc}
\includegraphics[width=0.26\textwidth]{fig/dt-f5} &
\includegraphics[width=0.26\textwidth]{fig/dt-f6} \\
Frame $i$ & Frame $i+1$ \\
\includegraphics[width=0.26\textwidth]{fig/dt-f7} &
\includegraphics[width=0.26\textwidth]{fig/dt-f8} \\
Frame $i+2$ & Frame $i+3$ \\
\end{tabular}
}
\caption{\label{fig::motiv}
Example of static tunnel detection in DhaA haloalkane dehalogenase and the example of a dynamic tunnel that evolves through four consecutive frames.
In the frame $i$, the active site (red sphere) is not fully connected with a protein surface, but only to an internal cavity.
The cavity disconnects from the active site in the frame $i+1$ and remains disconnected also in the next frame $i+2$.
In the fourth frame $i+3$, the cavity connects with the protein surface.
}
\end{figure}

@@

Alternatively, the tunnel detection in dynamic proteins can be formulated as path planning in 4D (i.e., 3D $\times$ time) configuration space assuming a spherical probe.
This space can be searched using sampling-based motion planning methods like Probabilistic Roadmaps (PRM)~\cite{kavrakiPRM} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} and their modifications.
In sampling-based motion planning, the configuration space is randomly sampled and the samples are classified as free or non-free using collision detection.
The collision detection is performed between the geometry of the robot (i.e., the probe) and the environment (the atoms of the protein).
The collision-free samples are stored into a roadmap which is searched using standard graph-search methods for a path.
The path in the roadmap then describes a motion in the workspace.

@@

In~\cite{guieysse2008structure,cortes2010simulating}, RRT is used to compute exit pathways for a flexible ligand leading from an active site to the surface of the protein.
To cope with many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.
In~\cite{cortesMD}, the pathway is found for the opposite direction, from the active site towards the protein surface.
The methods~\cite{guieysse2008structure,cortesMD} do not compute all tunnels explicitly, rather a single pathway.

@@

A possible solution is to estimate the location of narrow passages from the knowledge of the workspace and generate more samples
in the difficult areas.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling e.g.
around medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}, or even by shifting
uniformly generated samples towards the medial axis~\cite{amatoOBPRM} or its approaximation~\cite{hollemanMAPRM}.
Another approach is to estimate difficult regions online, e.g. based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
Generating random samples in difficult regions however does not ensure the construction of a better tree in the case of RRT-based methods.
As RRT builds the tree incrementally, the tree can connect to a region with increased probability of sampling only if the growth
of the tree is not blocked by obstacles.
Instead of increasing probability of sampling in all regions at once, e.g. along the whole medial axis, the tree has to be guided through the space to reach the regions gradually.
For example, the random samples are generated from multiple regions, that are switched according to the progress of the tree.
In~\cite{kardossRRTKK}, the samples are generated along several waypoints that are located close to narrow passages.
A more general approach is to define a guiding path in the workspace and sequentially increase the probability of sampling along
the path~\cite{vonasek2009rrt,denny2014marrt,denny2016dynamic}.
%The guiding path can be computed e.g. using medial axis of the worskpace~\cite{denny2014marrt,denny2016dynamic}.

@@

Molecular dynamics simulation is represented by a sequence of consecutive snapshots (frames), where each snapshot describes the position of all protein atoms in a specific time.
Proteins are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
The geometry of the protein in the $i-$th frame is represented by the union of all spheres $\SS^i \subset \R^3$.
The void space is detected using the spherical probe of radius $\probe$. 
All possible configurations $q=(x,y,z)$ (positions) of the probe in the $i$-th frame form the configuration space $\C^i$.
The probe is represented by a sphere $\Sprobe(q) \subset \R^3$ at a configuration $q$.
The void space (the collision-free region) $\CF^i \subseteq \C^i$ is formed by configurations, where $\Sprobe$ can be placed without any collision with atoms, i.e.,  $\CF^i = \{q \in \C^i | \Sprobe(q) \cap \SS^i = \emptyset\}$.
For each configuration $q\in\C^i$ we assume that its largest collision-free radius $r(q) \in \mathbf{R}, r(q)\ge 0$ can be computed, e.g., 
using collision detection.
The outer space  $\CG^i \subset \CF^i$ of the protein is defined by configurations, where a large sphere can be placed without any collision,
i.e., $\CG^i=\{q\in \C^i| \Sgprobe(q) \cap \SS^i = \emptyset \}$ where $\Sgprobe$ is the sphere of radius $\gprobe > \probe$ 
placed at the configuration $q$.
The exit points, where the tunnels leave the protein, are located in $\CG^i$.
To determine whether two configurations $q_1, q_2 \in \CF^i$ can be connected, a line segment from $q_1$ to $q_2$ is constructed
and discretized to several points, using $\varepsilon$ discretization step, and the collision detection is performed at all the points.
The two configurations can be connected if all intermediate points on the line are collision-free. 
The 3D Euclidean metric is used to measure distances in the configuration space.

The main purpose of the tunnel detection is the identification of tunnels, either static or dynamic, that connect the active site with the protein surface.
We assume that the position (configuration) of the active site $\qinit \in \CF^i$ in each snapshot is known in advance 
or it is known that the probe cannot be placed at the active site.
A static tunnel is a collision-free path leading from $\qinit \in \CF^i$ to an exit point $q \in \CG^i$, i.e., both active site
and the exit point are located in the same frame $i$.
A dynamic tunnel is a collision-free path leading from $\qinit \in \CF^i$ to an exit point $q \in \CG^j$, $j>i$ i.e., the active size and the
 exit point are located in different frames.

The approximated WVD diagram of the atoms is computed by representing each atom using 12 balls of equal radii and computing the 
ordinary Voronoi diagram.
Computing WVD this way is numerically more stable and it is used also in other related tools~\cite{caver3,yaffe2008}.
The vertices of this WVD are filtered out if their distance to an atom is less than the probe radius $\probe$ or if their
distance to the nearest atom is larger than $\gprobe$.
These filtered Voronoi vertices are stored in the set $\VV^i$.
The atoms located on the surface of the molecule can be identified using \mbox{$\alpha$-shape} with radius $\gprobe$.
Let $\SSA^i \subset \SS^i$ denote a set of spheres representing these atoms in snapshot $i$.

@@

To detect the void space, the collision-free region $\CF^i$ of the $i$-th frame is searched and represented as a tree structure of the collision-free configurations.
The detection starts from a given active site $\qinit \in \CF^i$ or from a set of already detected collision-free positions of the probe.
The second case is a result of the void space detection in the previous frame of the molecular dynamics.
To maintain the connectivity between the void space from the previous frame and the newly discovered one 
(i.e., from $\CF^i$ to $\CF^{i+1}$, the void space has to be explored incrementally without adding non-reachable regions, such as cavities.
Due to this reason, the void space cannot be automatically initialized using all Voronoi vertices $\VV^i$, though they are collision-free
at the radius $\probe$, because the Voronoi vertices can be also placed in the cavities.
The required incremental detection of the void space can be achieved by the exploration principle of the RRT planner, as RRT incrementally extends a tree of collision-free configurations.
The main loop of the algorithm for the void space detection is listed in Alg.~\ref{alg::main}.

@@

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
{\small
\caption{\label{alg::main}voidSpaceDetection}
\KwIn{Initial configuration tree $\T$,
    initial set of blocking spheres $\SB'$,
    position of active site $\qinit\in\CF^i$,
}
\KwData{goal-probe radius $\gprobe$,
    probe radius $\probe$, 
    number of iterations $\Imax$, 
    phase--switching parameter~$\alpha$,
    prob. $\gb$~of sampling randomly from~$\C^i$,
}
\KwOut{Configuration tree $\T$,
    list of tunnel exit-points $P$
}
\hrule
$\SB = \SB'$; //blocking atoms added during phase~1 \\
$P = \emptyset$; // list of tunnel exit-points \\
$phase=1$\;
$\VV^i = \VV^i \cup \{ \qinit \}$; // add active site position in order to reach it by the tree\\
updateCollisionDetection($\SB$)\;       
\For{$iteration = 1:\Imax$}{
    \If(// Voronoi-based sample --- Section~\ref{sec::vbg}){$phase=1$}{
        $\VVA^i$ = updateActiveVoronoiVertices($\T, \VV^i$)\;
        $\VVA^i$ = sort vertices according to the distance to the tree\; \nllabel{alg::main::d}
        \ForEach{$v \in \VVA^i$}{
            $\qnear$ = nearestNode($\T, v$)\;
            $\T,P', \SB'$ = expandTree($\T, \qnear, v$)\;
            $P = P \cup P'$\; \nllabel{alg::main::g1}
            $\SB = \SB \cup \SB'$\;
            updateCollisionDetection($\SB$)\; \nllabel{alg::main::a} \nllabel{alg::main::e}
        }
    }
    \If{$phase=1$}{ \nllabel{alg::main::f}
        $v =$ random point from $\VVA^i$\;
        $\qrand =$ random point $\sim N(v,r_v)$\; \nllabel{alg::main::gauss}
    }
    \If{$phase=2$ {\bf or} rand(0,1) $< p_c$}{
        $\qrand$ = random sample from $\C^i$\; \nllabel{alg::main::rand}
    }

@@

Besides the exploration of the void space, it is necessary to detect the exit points as well.
The exit points are configurations from $\CG^i$, i.e., they are collision-free at the radius $\gprobe$ and they define the end of the tunnel.
The detection of the exit points requires that the tree exits the protein.
A premature exit from the protein may lead to quick filling of the outer part of the protein and eventually even to entering the protein from outside. 
Such an exploration would lead to the detection of a smaller number of tunnels.
An example as depicted in Fig.~\ref{fig::blocking1}.

To ensure that the tree explores primarily the internal void space while allowing its growth outside the protein for the purpose of exit point detection, the algorithm is divided into two phases.
In the first phase, the growth of the tree outside the protein is blocked using a set of {\sl blocking spheres} $\SB$.
The blocking spheres are placed at configurations that are located close to the surface atoms $\SSA^i$.
After a configuration $q$ is added to the tree, the distance between $q$ and the surface atoms is computed.
If the distance is smaller than a predefined threshold $\dts$, $q$ can be assumed to be close to the surface and a new blocking sphere of radius $\gprobe$ is added to $\SB$.
By adding the blocking sphere at $q$, the tree cannot further expand not only from $q$, but also from the nearby nodes, which ensures,
   that the tree will not exit the protein in the first phase of the void space detection.

In the second phase, the blocking spheres are not used and they are removed from the collision detection (Alg.~\ref{alg::main}, line~\ref{alg::main::b}).
This allows the tree to explore also the void space outside the protein, which is necessary to find configurations representing the tunnel exit points.
The outer space is more likely reached from the nodes that could not be expanded in the first phase due to the blocking spheres.
As these nodes are located close to the surface atoms, this situation does not require many iterations to expand these nodes.
The second phase of the algorithm should therefore be shorter than the first one.
In Alg.~\ref{alg::main}, the second phase starts after $\alpha$ fraction of the iterations.

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{1pt}
\begin{tabular}{cccc}
\includegraphics[width=0.24\textwidth]{fig/blockingb1} &
\includegraphics[width=0.24\textwidth]{fig/blockingb2} & 
\includegraphics[width=0.24\textwidth]{fig/blockingb3} & 
\includegraphics[width=0.24\textwidth]{fig/blockingb4} \\
a & b & c & d \\
\end{tabular}  
}
\caption{\label{fig::blocking1}
    Example of the tree growth without the blocking spheres. 
    The initial configuration $\qinit$ (active site) is represented by the black circle.
    The tree starts to grow inside the protein (a) but it can grow outside as well (b).
    In region denoted by A, the tree grows back to the protein (c).
    Finally, two tunnels are detected (with the exit points $P_1$ and $P_2$), but no tunnel is found in the region $A$ (d).
}
\end{figure}

\begin{figure}
\centering
{
\renewcommand{\tabcolsep}{1pt}
\begin{tabular}{cccc}
\includegraphics[width=0.24\textwidth]{fig/blockingb5} &
\includegraphics[width=0.24\textwidth]{fig/blockingb6} &
\includegraphics[width=0.24\textwidth]{fig/blockingb7} & 
\includegraphics[width=0.24\textwidth]{fig/blockingb9} \\
a & b & c & d \\
\includegraphics[width=0.24\textwidth]{fig/blockingb10a} &
\includegraphics[width=0.24\textwidth]{fig/blockingb10b} & 
\includegraphics[width=0.24\textwidth]{fig/blockingb10c} & 
\includegraphics[width=0.24\textwidth]{fig/blockingb10d} \\
e & f & g & h \\
\end{tabular}
}
\caption{\label{fig::blocking2}
The effect of the blocking spheres. 
The surface atoms are depicted in brown.
After the tree reaches a surface atom, a new blocking sphere (red large circle) of size $\gprobe$ is added to the collision detection (a).
This forces the tree to explore only the internal void space (b,c).
In the second phase of the algorithm, the blocking spheres are removed and the tree can finally grow outside the protein in order
to detect exit points (d).
Figures e--h depict four detected tunnels.
}
\end{figure}

@@

The generation of the random samples directly influences the performance of the RRT-based planners.
In classic RRT, the random samples are drawn uniformly in the whole configuration space, which leads to the well known narrow passage problem.
The void space can be considered as a narrow passage, as the movement of the probe is constrained by many atoms.
The limitation is even higher for larger probes.
To boost the growth of the tree inside the narrow void space, the random samples can be generated around places that are believed to be a part
of the collision-free region $\CF^i$.
This region is approximated by the Voronoi vertices $\VV^i$ and the random samples can be generated around them.

However, the probability of sampling cannot be increased around all Voronoi vertices at once, because the growth of the tree towards the Voronoi vertices may be blocked by the obstacles (atoms)~\cite{vonasekphd}.
It is necessary to guide the growth of the tree by increasing the probability of sampling only around such Voronoi vertices that can be reached by the tree with a higher probability.
Such Voronoi vertices have to be located close to the tree.
Therefore, a subset $\VVA^i \subseteq \VV^i$ of the Voronoi vertices that are close to the tree is maintained (Alg.~\ref{alg::update}).
A Voronoi vertex $v \in \VV^i$ is considered as active if the distance to its nearest node in the tree is less than $\da$, but larger than
$\db$.
After the update of $\VVA^i$, the tree attempts to expand towards them (Alg.~\ref{alg::main}, line~\ref{alg::main::d}--\ref{alg::main::e}).
This expansion is not always successfull, especially for distant Voronoi vertices, but it helps the tree to explore
the void space.
